---
title: "Short recapitulation ggplot"
author: "Mark Klik & Misja Mikkers"
output: html_notebook
---

# Packages

```{r}
library(tidyverse)
```


# Short recapitulation ggplot

## Background and design of the ggplot2 package

* In 2005  Hadley Wickham created ggplot as part of his PhD.
* ggplot is an implementation of the 'Grammar of Graphics' (Wilkinson, 2005). 
* `ggplot2` makes it possible to link a certain _aesthetic_ of a graph to a column in a dataframe. Examples of are _Aesthetics_ :

    - color fill
    - line color
    - line size
    - size and shape of points
    - transparancy (alpha)
    - x-coordinate
    - y-coordinate
    - line type
    - etc

## small example

We will use the data-set of Americain presidents available of `R`:

    
```{r, }
df <- presidential
df
```


```{r}
ggplot(df) +
  geom_point(aes(start, name, size = start, colour = party)) +
  geom_point(aes(start, name, shape = factor(end - start)), size = 1.5) +
  labs(shape = "duration in days", size = "Starting year", colour = "Party") +
  xlab("Start presidency") +
  ylab("Name of the president") +
  geom_text(aes(start, name, label = substr(name, 2, 3)), nudge_y = 0.35)
```

In this graph we made the following links between the _aesthetics_ and the columns:

* color of the points is linked to the party of the president
* the x-as is linked to the starting date of the presidency
* the y-as is linked to the name of the president
* the shape in the point is linked to the number of days in office
* the letter is linked to the second and third letter of the name of the president



